Liveness-Based Garbage Collection for Lazy Languages

### 2015-06-06 paper/paper.tex
Updated example grammar
liveness

The  solution  is  to   also  treat  the  operations  mentioned  above
symbolically.  We  introduce three  new  symbols.   First, \bcar\  and
\bcdr\ are defined as:
\begin{align*}
 \bcar\sigma &\triangleq \{\alpha \mid 0\alpha \in \sigma\}\\
 \bcdr\sigma &\triangleq \{\alpha \mid 1\alpha \in \sigma\}
\end{align*}
and \cbar\ as:
\begin{align*}
 \cbar\sigma \triangleq\; & \emptyset~\mbox{if}~\sigma = \emptyset\\
                       & \{\epsilon\} ~\mbox{otherwise}
\end{align*}
Note that \cbar\ is idempotent. 
We can now rewrite the \CONS\ and the \CAR\ rules of $\mathit{ref}$ as:
\begin{align*}
&\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
= x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons},~
\mbox{and} \\
&\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          =   x.\cbar\sigma \cup x.0\sigma
\end{align*}
and the \Lfunonly\ rule
for \SIF\ as:
\begin{align*}
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) =
                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\uplus
        \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
        &[\,\mathit{ep} \mapsto  x.\cbar \sigma]
\end{align*}
The rules for \CDR, \PRIM\ and  \NULLQ\  are also modified in a manner
similar to \CAR.
The  defintions of \bcar\, \bcdr\ and \cbar\  induce   the  following
relation $\hookrightarrow$ over sets of access paths:
\begin{align*}
  &\sigma_1\bcar\sigma_2  \hookrightarrow
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
&\sigma_1\bcdr\sigma_2  \hookrightarrow
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
  \acdr\sigma_2',~\mbox{and}\\
&\sigma_1\cbar \sigma_2 \hookrightarrow \epsilon, \mbox{ if } \sigma \ne
  \emptyset 
\end{align*}
The reflexive  transitive closure of  $\hookrightarrow$ will
be   denoted    as   $\stackrel{*}{\hookrightarrow}$.    The
following proposition relates  the original and the modified
liveness rules.

@@

        \small
\begin{eqnarray*}
        \mathcal{L}(e_{\length},\sigma,\Lfonly) &=&
              [\;ep_1 \mapsto \lbrace \px.\cbar \sigma,
\xl.\cbar\sigma\rbrace ,~ep_2 \mapsto \lbrace \cred{
\pv.\sigma}  \rbrace,\\
 &&          \;\;  ep_3 \mapsto \lbrace \pz.\sigma,
                \py.\cbar\sigma,
\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma},\\
&&            \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\; 
\xl.\{ \acdr\Lf{\length}{\mbox{1}}{\cbar\sigma} \cup 
%%\;\;\;\;\;\;\; \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\\
\cred{\cbar\Lf{\length}{1}{\cbar\sigma}}\}\}
             \rbrack \\
            \cal{P_{\mbox{\tiny \length}}} &=&
                [\;\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace,  \
                 \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
                  \rbrace, \\
&&\;\;  \pi_3 \mapsto \lbrace ep_2
                  \rbrace ,  \pi_4 \mapsto \lbrace ep_2
                  \rbrace, \pi_5 \mapsto \lbrace  ep_3 \rbrace, \\
 &&  \;\;                        \pi_6 \mapsto \lbrace ep_3 \rbrace,
\pi_7 \mapsto
                   \lbrace ep_3\rbrace, \pi_8 \mapsto \lbrace ep_3
\rbrace~\rbrack\\
              \mathcal{L}(e_{\main},\sigma,\Lfonly) &=&
                [\;
                  ep_4 \mapsto \lbrace \cred {\pw.\sigma,
                                       \pc.\Lf{\length}{\mbox{1}}{\sigma},}\\
&& \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;
                                        \cred{\pb.\bcar\Lf{\length}{\mbox{1}}{\sigma},}
                                        \cred{\pa.\cbar\bcar\Lf{\length}{\mbox{1}}{\sigma}} \rbrace
                               ] \\
               \cal{P_{\mbox{\tiny \main}}} &=&
                  \lbrack~ \cred{\pi_9 \mapsto \lbrace ep_4 \rbrace,  \
                    \pi_{10} \mapsto \lbrace ep_4 \rbrace,  \
                   \pi_{11} \mapsto \lbrace ep_4 \rbrace,} \\
                   && \;\; \cred{\pi_{12} \mapsto \lbrace ep_4 \rbrace, \pi_{13} \mapsto \lbrace ep_4 \rbrace}~\rbrack
\end{eqnarray*}
    \end{minipage} \\
(a)&
(b)
    %% \end{ tabular}}
  \end{tabular}
%\end{picture}
\kern -3ex
%\vspace{0.5in}
\caption{(a) Example program and (b) its liveness
maps.}\label{fig:mot-example2}
\end{figure*}
%================================================================

@@

\subsection{Generating Liveness Equations}
The  program  in  Figure~\ref{fig:mot-example2}  serves as  a  running
example.   Given a  function $\mathit{f}$,  we now  describe  how to
generate    equations    for     the    demand    transformation
\Lfonly$_\mathit{f}$.   Starting with a  symbolic demand  $\sigma$, we
determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.  In  particular, we
consider $\Lanv{1}{x_i}$,  the  liveness of    the
$i^{\mbox{\footnotesize  th}}$
parameter $x_i$ at the  program point at the  beginning of
$e_{\mathit{f}}$ (assumed  to be  $\pi_1$). By the rule {\sc
  live-define}, this should be the
same as $\Lf{f}{i}{\sigma}$. Applying this to  \length, we have:
\begin{eqnarray*}
&& \Lanv{1}{\xl} = \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma}
  \cup \cbar\Lf{\length}{1}{\cbar\sigma}
\end{eqnarray*}
and  the only equation defining \Lfone{\length} is:
  \begin{eqnarray*}
   && \Lf{\length}{1}{\sigma}
    =  \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma} \cup \cbar\Lf{\length}{1}{\cbar\sigma}
\end{eqnarray*}

In general, the equations for  \Lfonly\ are recursive since, as in the
case  of  \length,  \Lan{1}\  may   have  been  defined  in  terms  of
\Lfone{\mathit{f}}.  However,  it is desirable  to have a  closed form
solution for $\Lfone{\mathit{f}}$.  As mentioned in \cite{asati14lgc},
each  of the liveness  rules modifies  a demand  by prefixing  it with
symbols  in the  alphabet  $\lbrace \acar,  \acdr,\bcar, \bcdr,  \cbar
\rbrace$, and therefore we can assume that $\Lf{f}{i}{\sigma}$ has the
closed form:
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} = \Df{f}{i}\sigma
\end{eqnarray}
where \Df{f}{i} are sets of strings over the alphabet mentioned above.
Substituting   the   guessed   form   in   the   equation   describing
\Lfonly$_{\mathit f}$, and factoring  out $\sigma$, we get an equation
for  \Df{f}{i} that  is  independent of  $\sigma$.   Any solution  for
\Df{f}{i}  yields a  solution for  \Lfonly$_{\mathit f}$.   Applied to
\Lfonly$_{\mathit \length}$, we get:
  \begin{eqnarray*}
&&  \Lf{\length }{1}{\sigma} = \Df{\length}{1}\sigma,~\mbox{and}\\
&&   \Df{\length}{1} = \cbar \cup \acdr\Df{\length}{1}\cbar 
       \cup \cbar\Df{\length}{1}{\cbar}
  \end{eqnarray*}

Note that this equation can also be viewed as a CFG with \{\acdr,
\cbar\} as terminal symbols and \Df{\length}{1} as the sole
non-terminal.

@@

Consider a function $g$ containing a  call to $f$ at a site $\pi$, say
$\pi\!\!:\!(\LET~x   \leftarrow   (f\,y_1\,\ldots\,y_n)~\IN~\pi_i:e)$.
Let the  demand on $g$  be $\sigma_g$ and,  based on this  demand, the
liveness of  $x$ at $\pi_i$ be  $\Lanv{i}{x}$.  By the  $\LET$ rule of
Figure~\ref{fig:live-judge},    the   call   at    $\pi$   contributes
$\Lanv{i}{x}$ to the demand $\sigma_f$. Let us denote the contribution
of a call  site $\pi$ in a  function $g$ to the overall  demand on the
function $f$  as $\deltacall{f}{\pi}{g}$. Assuming that  there are $k$
call sites to function $f$, $\pi^1$ (in function $g^1$) \ldots $\pi^k$
(in   function  $g^k$),  the   over-approximation  of   $\sigma_f$  is
$\deltacall{f}{\pi^1}{g^1}          \cup          \cdots          \cup
\deltacall{f}{\pi^k}{g^k}$.  The distinguished function \mainpgm\ is a
special case.   We assume  it is called  through a  printing mechanism
with   demand   $\sigma_\mainpgm   =  \{\acar,\acdr\}^\ast$   (denoted
$\sigma_{\!all}$).

For  the  running  example,   $\length$  has  calls  from  $\main$  at
$\pi_{12}$ and a recursive call at $\pi_6$.
So $\sigma_{\length} =
     \deltacall{\length}{\pi_{12}}{\main}  \cup 
\deltacall{\length}{\pi_6}{\length}$.
Filling in  the values gives:
\begin{eqnarray*}
\sigma_{\length}    &=&
 \sigma_{\!all}  ~\cup~{\cbar\sigma_{\mathit{\length}}}
\end{eqnarray*}
As examples, the liveness of \Lanv{1}{\pl} and \Lanv{9}{\pc} in terms
of  $\sigma_{\length}$ are:
\begin{align*}
\Lanv{1}{\pl} &= \cbar\sigma_{\length} \cup \acdr\Df{\length}{1}\cbar\sigma_{\length}
  \cup \cbar\Df{\length}{1}{\cbar\sigma_{\length}}
 \\
\Lanv{9}{\pc} &= \Df{\length}{1}\sigma_{\mathit{\!all}}
\end{align*}

In summary, the equations generated during liveness analysis are:
\begin{enumerate}
\item   For    each   function   $\mathit{f}$,    equations   defining
\Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item  For  each   function  $\mathit{f}$, an  equation  defining
the summary demand  $\sigma_{\mathit f}$ on $e_f$.
\item For each function $\mathit{f}$ (including $\mainpgm$ for
  $e_\mainpgm$) 
an equation defining
  liveness at each GC point of $e_{\mathit f}$.
\end{enumerate}
\subsection{Solving liveness equations---the grammar
interpretation}\label{sec:grammar-formulation}
The  equations above can now be re-interpreted as a
context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
\bcar, \bcdr, \cbar\rbrace$.  Let  \var{$X$} denote the non-terminal
for
a variable  $X$ occurring on  the LHS of the  equations generated
from the analysis.  We can  think of the resulting productions as
being associated with several  grammars, one for each non-terminal
\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
grammar    for   \var{\Lanv{1}{\xl}}   comprises    the   following
productions:
 \begin{eqnarray*}
\var{\Lanv{1}{\xl}}  &\rightarrow&   \cbar\sigma_{\length} \mid \acdr \var{\Df
  {\length}{1}}{\cbar\sigma_{\length}} \mid
  \cbar\var{\Df{\length}{1}}{\cbar\sigma_{\length}} \\
  \var{\Df{\length}{1}} &\rightarrow& \cbar \mid
  \acdr\var{\Df{\length}{1}}\cbar 
       \mid \cbar\var{\Df{\length}{1}}{\cbar}\\
\langle {\sigma_{\length}} \rangle
&\rightarrow&
\cred{\epsilon}  \mid \cbar\langle{\sigma_{\length}}\rangle
\end{eqnarray*}
Other equations can be converted similarly.  The language generated by
\var{\Lanv{i}{x}},  denoted $\mathscr{L}(\var{\Lanv{i}{x}}) $,  is the
desired  solution of  \Lanv{i}{x}.  However,  recall from  our earlier
discussion that the decision problem  that we are interested in during
garbage collection at a program point $\pi_i$ is:
\begin{quote}
Let  $x.\alpha$ be  a forward  access  path consisting  only of  edges
\acar\    and   \acdr\    (but    not   \bcar, \bcdr\ or \cbar).     Let
$\mathscr{L}(\var{\Lanv{i}{x}}) \stackrel{*}{\hookrightarrow} \sigma$,
where $\sigma$ consists  of forward paths only. Then  does $\alpha \in
\sigma$?
\end{quote}

@@

\begin{definition}\label{def:specialgrammar} 
Consider a form of  grammar $(N,T, p_1\cup p_2,S)$ in which $N$ is a set of
non-terminals, $T = \{\acar, \acdr, \bcar, \bcdr, \cbar, \$\}$, $p_1$
is a set of context-free production that contains the distinguished
production $S \rightarrow \alpha\$$, $\alpha$ is a string of grammar
symbols that does not contain $S$, and $p_2$ is the fixed set of
unrestricted productions $\bcar\acar \rightarrow \epsilon$,
$\bcdr\acdr \rightarrow \epsilon$, $\cbar\acar \rightarrow \cbar$,
$\cbar\acdr \rightarrow \cbar$, and $\cbar\$ \rightarrow
\epsilon$. 
\end{definition}

@@

%------------------------------------------------------------%
\subsubsection{Approximating CFGs using NFAs.}
\label{sec:NFA-approx}
 We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
 approximate   a  CFG   to   a  {\em   strongly  regular\/}   grammar.
 \comment{Rewrite  this after  constructing the  example: We  show the
   grammar for ????? after  the Mohri-Nederhof transformation. We pick
   the only  production that  is affected by  the transformation---the
   production    for     \Df{\append}{1}.     The    production    for
   \Uf{\append}{1},  while recursive, is  already in  strongly regular
   form and is therefore unaffected by the transformation.
 \begin{eqnarray*}
  \var{\Df{\append}{1}}
  &\rightarrow&               \acar\bcar \var{\Df{\append}{1}}'  \mid
  \acdr\var{\Df{\append}{1}} \\
\var{\Df{\append}{1}}'  &\rightarrow&   \bcdr\var{\Df{\append}{1}}'
\mid \epsilon
\end{eqnarray*}
The  languages
generated for  \var{\Df{\append}{1}} in  the original grammar  and the
new       grammar      are       \acdr$^i$\acar\bcar\bcdr$^i$      and
\acdr$^*$\acar\bcar\bcdr$^*$, showing a  loss of  precision.

Once the NFA's are created, we need to remove the transitions on
$\cbar$. This edge is replaced by an $\epsilon$ transition and the
$\cbar$ edge is deleted. If the original grammar had a non-null
$\sigma$ then the NFA reaches a final state throught this path
otherwise it does not reach a final state.}

%------------------------------------------------------------%
\subsubsection{Transforming NFAs to Accept Forward Paths:}
%\label{sec:NFA-elim}
The   strongly  regular   CFGs  obtained   after   the  Mohri-Nederhof
transformation are first converted into NFAs.  The algorithm described
in  Figure~\ref{algo:simplify-NFA} converts  an NFA  \ntbar\ to  a NFA
\nt\    such    that   $\lang{\ntbar}    \stackrel{*}{\hookrightarrow}
\lang{\nt}$, where {\nt} accepts forward paths only.  \comment{Rewrite
  the algorithm to handle \cbar  and update the example automata} Thus
\nt\ can be used to check membership of forward paths.

@@

In this  section we describe in  detail the process  of liveness based
garbage collection.  During  the execution of a program  if it runs of
out  of memory,  execution is  stopped  and the  garbage collector  is
invoked. At this point all the references on the stack are checked for
liveness and only live references are copied and the program continues
execution.  The main  complexity in our algorithm is  the way closures
are  handled. The  following sub-sections  show two  different  ways of
handling closures along with their pros and cons.
\subsection{A simple LGC Algorithm}
\cred{Introduce terminology}
We introduce certain terminology that we use in our discussion of the 
garbage collection algorithms. A reference is a field which holds the 
address of a heap location. Root set refers to the set of all references 
that are on the program stack during garbage collection.
A heap cell can either contain an atomic value, a \CONS\ cell or a closure. A closure 
has fields to identify the closure type and 2 arguments. In case of function calls 
with more than 2 arguments we use currying (one of the argument fields will point to another
 heap location containing the next argument and so on). By reachability we mean all heap 
locations that are reachable traversing the argument fields of a given reference.
% Abstract description of the GC algorithm
\begin{verbatim}
proceudre lgc()
 FOREACH reference in root set
 DO
   IF reference is live
     copy(cell)
   ENDIF
 DONE

>>>>

The  solution  is  to   also  treat  the  operations  mentioned  above
symbolically.  We  introduce three  new  symbols.   First, \bcar\  and
\bcdr\ are defined as:
\begin{align*}
 \bcar\sigma &\triangleq \{\alpha \mid 0\alpha \in \sigma\}\\
 \bcdr\sigma &\triangleq \{\alpha \mid 1\alpha \in \sigma\}
\end{align*}
and \clazy\ as:
\begin{align*}
 \clazy\sigma \triangleq\; & \emptyset~\mbox{if}~\sigma = \emptyset\\
                       & \{\epsilon\} ~\mbox{otherwise}
\end{align*}
Note that \clazy\ is idempotent. 
We can now rewrite the \CONS\ and the \CAR\ rules of $\mathit{ref}$ as:
\begin{align*}
&\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
= x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons},~
\mbox{and} \\
&\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          =   x.\clazy\sigma \cup x.0\sigma
\end{align*}
and the \Lfunonly\ rule
for \SIF\ as:
\begin{align*}
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) =
                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\uplus
        \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
        &[\,\mathit{ep} \mapsto  x.\clazy \sigma]
\end{align*}
The rules for \CDR, \PRIM\ and  \NULLQ\  are also modified in a manner
similar to \CAR.
The  defintions of \bcar\, \bcdr\ and \clazy\  induce   the  following
relation $\hookrightarrow$ over sets of access paths:
\begin{align*}
  &\sigma_1\bcar\sigma_2  \hookrightarrow
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
&\sigma_1\bcdr\sigma_2  \hookrightarrow
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
  \acdr\sigma_2',~\mbox{and}\\
&\sigma_1\clazy \sigma_2 \hookrightarrow \epsilon, \mbox{ if } \sigma \ne
  \emptyset 
\end{align*}
The reflexive  transitive closure of  $\hookrightarrow$ will
be   denoted    as   $\stackrel{*}{\hookrightarrow}$.    The
following proposition relates  the original and the modified
liveness rules.

@@

        \small
\begin{eqnarray*}
        \mathcal{L}(e_{\length},\sigma,\Lfonly) &=&
              [\;ep_1 \mapsto \lbrace \px.\clazy \sigma,
\xl.\clazy\sigma\rbrace ,~ep_2 \mapsto \lbrace \cred{
\pv.\sigma}  \rbrace,\\
 &&          \;\;  ep_3 \mapsto \lbrace \pz.\sigma,
                \py.\clazy\sigma,
\pu.\Lf{\length}{\mbox{1}}{\clazy\sigma},\\
&&            \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\; 
\xl.\{ \acdr\Lf{\length}{\mbox{1}}{\clazy\sigma} \cup 
%%\;\;\;\;\;\;\; \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\\
\cred{\clazy\Lf{\length}{1}{\clazy\sigma}}\}\}
             \rbrack \\
            \cal{P_{\mbox{\tiny \length}}} &=&
                [\;\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace,  \
                 \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
                  \rbrace, \\
&&\;\;  \pi_3 \mapsto \lbrace ep_2
                  \rbrace ,  \pi_4 \mapsto \lbrace ep_2
                  \rbrace, \pi_5 \mapsto \lbrace  ep_3 \rbrace, \\
 &&  \;\;                        \pi_6 \mapsto \lbrace ep_3 \rbrace,
\pi_7 \mapsto
                   \lbrace ep_3\rbrace, \pi_8 \mapsto \lbrace ep_3
\rbrace~\rbrack\\
              \mathcal{L}(e_{\main},\sigma,\Lfonly) &=&
                [\;
                  ep_4 \mapsto \lbrace \cred {\pw.\sigma,
                                       \pc.\Lf{\length}{\mbox{1}}{\sigma},}\\
&& \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;
                                        \cred{\pb.\bcar\Lf{\length}{\mbox{1}}{\sigma},}
                                        \cred{\pa.\clazy\bcar\Lf{\length}{\mbox{1}}{\sigma}} \rbrace
                               ] \\
               \cal{P_{\mbox{\tiny \main}}} &=&
                  \lbrack~ \cred{\pi_9 \mapsto \lbrace ep_4 \rbrace,  \
                    \pi_{10} \mapsto \lbrace ep_4 \rbrace,  \
                   \pi_{11} \mapsto \lbrace ep_4 \rbrace,} \\
                   && \;\; \cred{\pi_{12} \mapsto \lbrace ep_4 \rbrace, \pi_{13} \mapsto \lbrace ep_4 \rbrace}~\rbrack
\end{eqnarray*}
    \end{minipage} \\
(a)&
(b)
    %% \end{ tabular}}
  \end{tabular}
%\end{picture}
\kern -3ex
%\vspace{0.5in}
\caption{(a) Example program and (b) its liveness
maps.}\label{fig:mot-example2}
\end{figure*}
%================================================================

@@

\subsection{Generating Liveness Equations}
The  program  in  Figure~\ref{fig:mot-example2}  serves as  a  running
example.   Given a  function $\mathit{f}$,  we now  describe  how to
generate    equations    for     the    demand    transformation
\Lfonly$_\mathit{f}$.   Starting with a  symbolic demand  $\sigma$, we
determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.  In  particular, we
consider $\Lanv{1}{x_i}$,  the  liveness of    the
$i^{\mbox{\footnotesize  th}}$
parameter $x_i$ at the  program point at the  beginning of
$e_{\mathit{f}}$ (assumed  to be  $\pi_1$). By the rule {\sc
  live-define}, this should be the
same as $\Lf{f}{i}{\sigma}$. Applying this to  \length, we have:
\begin{eqnarray*}
&& \Lanv{1}{\xl} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
  \cup \clazy\Lf{\length}{1}{\clazy\sigma}
\end{eqnarray*}
and  the only equation defining \Lfone{\length} is:
  \begin{eqnarray*}
   && \Lf{\length}{1}{\sigma}
    =  \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma} \cup \clazy\Lf{\length}{1}{\clazy\sigma}
\end{eqnarray*}

In general, the equations for  \Lfonly\ are recursive since, as in the
case  of  \length,  \Lan{1}\  may   have  been  defined  in  terms  of
\Lfone{\mathit{f}}.  However,  it is desirable  to have a  closed form
solution for $\Lfone{\mathit{f}}$.  As mentioned in \cite{asati14lgc},
each  of the liveness  rules modifies  a demand  by prefixing  it with
symbols  in the  alphabet  $\lbrace \acar,  \acdr,\bcar, \bcdr,  \clazy
\rbrace$, and therefore we can assume that $\Lf{f}{i}{\sigma}$ has the
closed form:
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} = \Df{f}{i}\sigma
\end{eqnarray}
where \Df{f}{i} are sets of strings over the alphabet mentioned above.
Substituting   the   guessed   form   in   the   equation   describing
\Lfonly$_{\mathit f}$, and factoring  out $\sigma$, we get an equation
for  \Df{f}{i} that  is  independent of  $\sigma$.   Any solution  for
\Df{f}{i}  yields a  solution for  \Lfonly$_{\mathit f}$.   Applied to
\Lfonly$_{\mathit \length}$, we get:
  \begin{eqnarray*}
&&  \Lf{\length }{1}{\sigma} = \Df{\length}{1}\sigma,~\mbox{and}\\
&&   \Df{\length}{1} = \clazy \cup \acdr\Df{\length}{1}\clazy 
       \cup \clazy\Df{\length}{1}{\clazy}
  \end{eqnarray*}

Note that this equation can also be viewed as a CFG with \{\acdr,
\clazy\} as terminal symbols and \Df{\length}{1} as the sole
non-terminal.

@@

Consider a function $g$ containing a  call to $f$ at a site $\pi$, say
$\pi\!\!:\!(\LET~x   \leftarrow   (f\,y_1\,\ldots\,y_n)~\IN~\pi_i:e)$.
Let the  demand on $g$  be $\sigma_g$ and,  based on this  demand, the
liveness of  $x$ at $\pi_i$ be  $\Lanv{i}{x}$.  By the  $\LET$ rule of
Figure~\ref{fig:live-judge},    the   call   at    $\pi$   contributes
$\Lanv{i}{x}$ to the demand $\sigma_f$. Let us denote the contribution
of a call  site $\pi$ in a  function $g$ to the overall  demand on the
function $f$  as $\deltacall{f}{\pi}{g}$. Assuming that  there are $k$
call sites to function $f$, $\pi^1$ (in function $g^1$) \ldots $\pi^k$
(in   function  $g^k$),  the   over-approximation  of   $\sigma_f$  is
$\deltacall{f}{\pi^1}{g^1}          \cup          \cdots          \cup
\deltacall{f}{\pi^k}{g^k}$.  The distinguished function \mainpgm\ is a
special case.   We assume  it is called  through a  printing mechanism
with   demand   $\sigma_\mainpgm   =  \{\acar,\acdr\}^\ast$ (denoted
$\sigma_{\!all}$) if \mainpgm\ returns a structure and $\epsilon$ if
it returns a base value.

For  the  running  example,   $\length$  has  calls  from  $\main$  at
$\pi_{12}$ and a recursive call at $\pi_6$.
So $\sigma_{\length} =
     \deltacall{\length}{\pi_{12}}{\main}  \cup 
\deltacall{\length}{\pi_6}{\length}$.
Filling in  the values gives:
\begin{eqnarray*}
\sigma_{\length}    &=&
 \{\epsilon\}  ~\cup~{\clazy\sigma_{\mathit{\length}}}
\end{eqnarray*}
As examples, the liveness of \Lanv{1}{\pl} and \Lanv{9}{\pa} in terms
of  $\sigma_{\length}$ are:
\begin{align*}
\Lanv{1}{\pl} &= (\clazy \cup \acdr\Df{\length}{1}\clazy
  \cup \clazy\Df{\length}{1}{\clazy})\sigma_{\mathit{\length}}
 \\
\Lanv{9}{\pa} &= \clazy \bcar \Df{\length}{1}\epsilon
\end{align*}

In summary, the equations generated during liveness analysis are:
\begin{enumerate}
\item   For    each   function   $\mathit{f}$,    equations   defining
\Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item  For  each   function  $\mathit{f}$, an  equation  defining
the summary demand  $\sigma_{\mathit f}$ on $e_f$.
\item For each function $\mathit{f}$ (including $\mainpgm$ for
  $e_\mainpgm$) 
an equation defining
  liveness at each GC point of $e_{\mathit f}$.
\end{enumerate}
\subsection{Solving liveness equations---the grammar
interpretation}\label{sec:grammar-formulation}
The  equations above can now be re-interpreted as a
context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
\bcar, \bcdr, \clazy\rbrace$.  Let  \var{$X$} denote the non-terminal
for
a variable  $X$ occurring on  the LHS of the  equations generated
from the analysis.  We can  think of the resulting productions as
being associated with several  grammars, one for each non-terminal
\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
grammar    for   \var{\Lanv{1}{\xl}}   comprises    the   following
productions:
 \begin{eqnarray*}
\var{\Lanv{1}{\xl}}  &\rightarrow&   \clazy\sigma_{\length} \mid \acdr \var{\Df
  {\length}{1}}{\clazy\sigma_{\length}} \mid
  \clazy\var{\Df{\length}{1}}{\clazy\sigma_{\length}} \\
  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
  \acdr\var{\Df{\length}{1}}\clazy 
       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
\langle {\sigma_{\length}} \rangle
&\rightarrow&
\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle
\end{eqnarray*}
Other equations can be converted similarly.  The language generated by
\var{\Lanv{i}{x}},  denoted $\mathscr{L}(\var{\Lanv{i}{x}}) $,  is the
desired  solution of  \Lanv{i}{x}.  However,  recall from  our earlier
discussion that the decision problem  that we are interested in during
garbage collection at a program point $\pi_i$ is:
\begin{quote}
Let  $x.\alpha$ be  a forward  access  path consisting  only of  edges
\acar\    and   \acdr\    (but    not   \bcar, \bcdr\ or \clazy).     Let
$\mathscr{L}(\var{\Lanv{i}{x}}) \stackrel{*}{\hookrightarrow} \sigma$,
where $\sigma$ consists  of forward paths only. Then  does $\alpha \in
\sigma$?
\end{quote}

@@

\begin{definition}\label{def:specialgrammar} 
Consider a form of  grammar $(N,T, p_1\cup p_2,S)$ in which $N$ is a set of
non-terminals, $T = \{\acar, \acdr, \bcar, \bcdr, \clazy, \$\}$, $p_1$
is a set of context-free production that contains the distinguished
production $S \rightarrow \alpha\$$, $\alpha$ is a string of grammar
symbols that does not contain $S$, and $p_2$ is the fixed set of
unrestricted productions $\bcar\acar \rightarrow \epsilon$,
$\bcdr\acdr \rightarrow \epsilon$, $\clazy\acar \rightarrow \clazy$,
$\clazy\acdr \rightarrow \clazy$, and $\clazy\$ \rightarrow
\epsilon$. 
\end{definition}

@@

%------------------------------------------------------------%
\subsubsection{From CFGs to GC-ready DFAs}
\label{sec:NFA-approx}
 We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
 approximate a CFG to a {\em strongly regular\/} grammar.  The grammar
 fragment  that  defines  the  non-terminal  $\var{\Df{\length}{1}}$
   after the Mohri-Nederhof transformation becomes:

 \begin{eqnarray*} 
   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
   \acdr\var{\Df{\length}{1}}
   \mid \clazy\var{\Df{\length}{1}}\\
   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid \epsilon \\
   \langle {\sigma_{\length}} \rangle
   &\rightarrow&
   \epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle
 \end{eqnarray*}

The grammar after the Mohri-Nederhoff transformation is converted into
a  set of  NFAs,  one for  each  $\var{\Lanv{i}{x}}$. We  now apply  the
$\hookrightarrow$ simplification  by repeatedly introducing $\epsilon$
edges to  bypass a pair  of consecutive edges labelled  \bcar\acar\ or
\bcdr\acdr  and  deriving  the  $\epsilon$-closure  of  the  resulting
automaton.   The   process  is  continued   till  a  fixed   point  is
reached.  Finally we  replace  the $\clazy$  edges  by $\epsilon$  and
determinize the automaton thus formed.

%------------------------------------------------------------%
\subsubsection{Transforming NFAs to Accept Forward Paths:}
%\label{sec:NFA-elim}
The   strongly  regular   CFGs  obtained   after   the  Mohri-Nederhof
transformation are first converted into NFAs.  The algorithm described
in  Figure~\ref{algo:simplify-NFA} converts  an NFA  \ntbar\ to  a NFA
\nt\    such    that   $\lang{\ntbar}    \stackrel{*}{\hookrightarrow}
\lang{\nt}$, where {\nt} accepts forward paths only.  \comment{Rewrite
  the algorithm to handle \clazy\  and update the example automata} Thus
\nt\ can be used to check membership of forward paths.

@@

In this  section we describe in  detail the process  of liveness based
garbage collection.  During  the execution of a program  if it runs of
out  of memory,  execution is  stopped  and the  garbage collector  is
invoked. At this point all the references on the stack are checked for
liveness and only live references are copied and the program continues
execution.  The main  complexity in our algorithm is  the way closures
are  handled. The  following sub-sections  show two  different  ways of
handling closures along with their pros and cons.
\subsection{A simple LGC Algorithm}
\cred{Introduce representation}
We introduce certain terminology that we use in our discussion of the 
garbage collection algorithms. A reference is a field which holds the 
address of a heap location. Root set refers to the set of all references 
that are on the program stack during garbage collection.
A heap cell can either contain an atomic value, a \CONS\ cell or a closure. A closure 
has fields to identify the closure type and 2 arguments. In case of function calls 
with more than 2 arguments we use currying (one of the argument fields will point to another
 heap location containing the next argument and so on). By reachability we mean all heap 
locations that are reachable traversing the argument fields of a given reference.
% Abstract description of the GC algorithm
\begin{verbatim}
proceudre lgc()
 FOREACH reference in root set
 DO
   IF reference is live
     copy(cell)
   ENDIF
 DONE
